Evaluation of Pre-Trained Models for Translation in Different Languages

### 2024-07-29 paper/paper.tex
updated code
updated, added opus

\title{Evaluation of Machine Translation in Languages}
\author{Stefan Liemawan Adji}
\date{\today}

@@

According to Ethnologue \cite{ethnologue-2024}, 7,164 languages currently exist and in use today, with 40\% of them considered endangered.

As of July 2024, 243 languages and languages varities are supported by Google Translate (according to Wikipedia \cite{wikipedia-google-translate})

Machine translation (MT) can be traced down to as early as 1949 \cite{weaver-1999}, recently growing as a popular research field within Natural Language Processing (NLP).

Statistical machine translation, which previously dominated MT research for many years with its reliance on various count-based models, has largely been surpassed by neural machine translation (NMT) \cite{stahlberg-2020-nmt-review}. Nowadays, Large Language Models (LLMs) such as mBart \cite{liu-2020-mbart}, M2M-100 \cite{fan-2020-m2m100}, NLLB-200 \cite{nllb200}, T5 \cite{raffel-2023-t5}, and GPT-4 \cite{openai-2024-gpt4-technical-report} dominated the field of machine translation. DeepL \cite{deepl} currently proclaimed itself as 'The world most accurate translator'.

Through simple experimentation and analysis, this paper aims to evaluate existing techniques across different languages
and contribute valuable insights into the current state of machine translation technology. By comparing metrics among languages, I hope to inform future researchers of the effectiveness of machine translation on each individual languages that are included in this paper.

\section{Related Work}

As is the case of most fields within Natural Language Processing (NLP), the current state-of-the-art (SOTA) in machine translation (MT) is characterised by the significant advancements brought by large language models (LLMs) like OpenAI's GPT-4. These models have surpassed traditional neural machine translation (NMT) systems in several areas, offering more nuanced and accurate translations, especially in handling context and idiomatic expressions.

\section{Dataset}

\subsection{Overview}

@@

@@

Sentences in English are downloaded, 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website). Then for each langauges, download sentence pairs compared to English. Merge every language sentences into one big dataframe, only keep where sentences exist for every language

>>>>

\title{Evaluation of Pre-Trained Models for Translation in Different Languages}
\author{Stefan Liemawan Adji}
\date{\today}

@@

According to Ethnologue \cite{ethnologue-2024}, 7,164 languages currently exist and in use today, with 40\% of them considered endangered. As of July 2024, 243 languages are supported by Google Translate (according to Wikipedia \cite{wikipedia-google-translate}). In modern times, the need for translation services has surged due to the growing exchange of information across different regions that speak various languages \cite{okpor-2014-machine-ta}.

Machine translation (MT) is the task of automatically translating from one language to another. This can be done through text or audio. It can be traced back to 1949 \cite{weaver-1999}, with the first public demonstration of an MT system on January 7, 1954, in collaboration with IBM, where 49 Russian sentences were translated into English using a limited vocabulary of 250 words and 6 grammar rules \cite{hutchins-2006-first-mt}. However, over the next several decades, growth were limited for machine translation, with 1956-1966 considered the decade of high expectation and disillusion, and 1967-1976 dubbed 'the quiet decade' \cite{hutchins-2001-mt-50-years}. Then in 1989, the dominance of the rule-based approach has been challenged by the rise of new methods and strategies, collectively referred to as ‘corpus-based’ methods (data-driven) \cite{hutchins-1994-research-methods-mt,hutchins-1998-development-mt}. Subsequently, statistics-based approaches for MT re-emerged, bolstered by the recent success of probabilistic techniques in speech recognition. Statistical machine translation \cite{lopez-2008-smt} dominated the domain between late 1990s through the early 2010s, before largely being surpassed by neural machine translation (NMT) \cite{cho-2014-properties,sutskever-2014-seq2seq}. Finally, since the introduction of Transformers in 2017 \cite{vaswani-2017-attention}, Natural Language Processing (NLP) and machine translation in particular reached a giant milestone. The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, and T5 \cite{raffel-2023-t5}, which revolutionised both MT and the whole field of NLP. Soon after, several pre-trained models that are specifically designed for machine translation emerged, namely mBART \cite{liu-2020-mbart}, mT5 \cite{xue-2021-mt5}, and NLLB \cite{nllb200}.

Since LLMs require vast amounts of training data, the availability of parallel corpora is crucial for building effective models. The lack of extensive parallel corpora, especially for low-resource languages, leads to suboptimal performance in NMT techniques compared to their high-resource counterparts \cite{ranathunga-2023-nmt-low-res}.

OPUS \cite{opus} is a comprehensive collection of open-source parallel corpora used extensively in the field of machine translation (MT). It includes corpora for 744 languages and contains over 1,210 different datasets, amassing a total of 45,945,946,108 sentence pairs. Tatoeba \cite{tatoeba} is another prominent resource in the field of MT and NLP, known for its extensive collection of translated sentences. As of July 2024, it contains 12,186,207 sentences over 423 supported languages, growing daily through volunteer contributions.

pre-trained models...

Through simple experimentation and analysis, this paper aims to evaluate existing pre-trained models across different languages
and contribute valuable insights into the current state of machine translation technology.

\section{Literature Review}

\subsection{Neural Machine Translation (NMT)}

The encoder-decoder approach \cite{cho-2014-properties} remains as the foundation architecture for many sequence-to-sequence models in natural language processing, especially machine translation.

\subsection{One-to-one Models}

One-to-One Translation refers to a translation approach where a model is specifically trained to translate between one source language and one target language. This setup is characterised by having a dedicated model for each unique language pair.

The OPUS-MT by Helsinki-NLP \cite{tiedemann-2020-opus-mt} is a good example of one-to-one translation project, they provide over 1,000 pre-trained translation models for many language pairs, allowing users to perform high-quality translations without needing to train models from scratch.

OpenNMT \cite{klein-etal-2017-opennmt} is another framework that allows for one-to-one translation, although their pre-trained models only support English and German languages.

\subsection{Many-to-many Models}

With the rise of LLMs and pre-trained language models, multilingual machine translation models have received popularity, particularly many-to-many translation.

Multilingual neural machine translation (NMT) allows for the training of a single model capable of translating between multiple source and target languages \cite{aharoni-2019-massively-multilingual}.

Many-to-many translation refers to the capability of a model to translate text between multiple language pairs bidirectionally. This means the model can handle translations from any of the supported source languages to any of the supported target languages, not limited to a single source or target language.

mBart \cite{liu-2020-mbart} (50 languages), NLLB-200 \cite{nllb200} (200 language), PolyLM \cite{wei-2023-polylm} (32 languages), and mt5 \cite{xue-2021-mt5} (101 languages) are some of the most popular many-to-many models.

\subsection{Sequence-to-Sequence Models}

GPT \cite{openai-2024-gpt4}

mBERT \cite{devlin-2019-bert} is the multilingual version of BERT, introduced in the same paper, trained on 104 different languages. However, the model been shown to suffer on low-resource languages \cite{wu-2020-mbert-are-all}.

T5 \cite{raffel-2023-t5} is a Text-to-Text Transfer Transformer model trained on English.

MistralAI \cite{jiang-2023-mistral-7b}

\section{Experiments}

\subsection{Dataset}

@@

Tatoeba is chosen due to its sentence having more daily life sentences...?

@@

Sentences in English are downloaded, 1,898,494 sentences (it is unclear why it is less than the number stated in the Tatoeba website). Then for each languages, download sentence pairs compared to English. Merge every language sentences into one big dataframe, only keep where sentences exist for every language
